---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# kosmic

<!-- badges: start -->
<!-- badges: end -->

This repository contains R bindings for Kosmic, a C++ library that can estimate reference intervals from routinely collected laboratory data. It does this by fitting a normal distribution to the central part of the observed data. The [original C++ library](https://gitlab.miracum.org/kosmic/kosmic) was written by Jakob Zierk and others.

## Installation

The R bindings for Kosmic are still under development. You can install the development version from GitHub.

``` r
# install.packages("devtools")
devtools::install_github("divinenephron/kosmic-r")
```

## Example

Here a reference interval is estimated for the simulated `haemoglobin` dataset which contains randomly generated physiological results and contamination from randomly generated pathological result. The true reference interval for this data is 12.0-16.0 g/dL, and the estimates from Kosmic are close.

```{r example}
library(kosmic)
k <- kosmic(haemoglobin$result, decimals = 1)
plot(k)
```

```{r}
summary(k)
```

